---
output: github_document
editor_options: 
  chunk_output_type: console
---


```{r setup, echo=FALSE, warning=FALSE, purl=FALSE, message=FALSE}
options(repos="http://cran.rstudio.com/")
opts_chunk$set(tidy=FALSE)
```
 
# Demo: Lake morphometry in R

## Lesson Outline
- [New script](#new-script)
- [Required packages](#required-packages)
- [Revisit our data](#revisit-our-data)
- [Get elevation data](#get-elevation-data)
- [Calculate Trout Lake morphometry](#calculate-trout-lake-morphometry)

## New script
We still should be working in the "rspatial_workshop" project we created in the map demo.  For this demo, we will create another new script.  

1. Go to File: New File: R Script.  
2. When that empty new file opens up, save it as "lakemorpho_demo.R"

## Required packages
```{r message=FALSE}
library(dplyr)
library(sf)
library(lakemorpho)
library(elevatr)
```

## Revisit our data
Just to make sure we have everything loaded up again and working


```{r}
ntl_lakes <- st_read("nhld_study_lakes.shp")
trout_lake <- ntl_lakes %>%
  filter(LAKE_NAME == "Trout Lake")
```




